---
title: "Whole root, half root, and strain models"
author: "Rebecca Batstone"
date: "`r format(Sys.Date())`"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## load packages

```{r packages}
library("tidyverse") ## includes ggplot2, dplyr, readr, stringr
library("reshape2") ## for melting datasheets
library("car") ## Anova function
library("cowplot") # paneled graphs
library("lme4") ## mixed effects models
library("knitr") ## knit analyses files
library("multcomp") ## post-hoc comps
library("multcompView") ## summarize key variables
library("fitdistrplus") ## determine best prob distributions
library("logspline") ## visualize prob dists
library("DHARMa") ## residual diagnostics for GLMMs
library("RVAideMemoire") ## residual diagnotics
library("emmeans") ## post-hoc contrasts, lsmeans
library("lattice") ## graphs

#model convergence packages
library("numDeriv")
# library("RCurl") ## to source() from Github
library("RColorBrewer") ## graph colors

source("./overdisp.R")
```

## load data

```{r load_data}
load("./prepped_data/plants.Rdata")
load("./prepped_data/plants_long.Rdata")
load("./prepped_data/plants_long_strain.Rdata")
```

## Set contrasts

https://rstudio-pubs-static.s3.amazonaws.com/65059_586f394d8eb84f84b1baaf56ffb6b47f.html

"By default, R uses traditional dummy coding (called “treatment contrasts” in R) for any non-ordered factors, and polynomial trend contrasts for any ordered factors. That works out well if you intend to look at regression coefficients"

"Note that traditional dummy coding is fine for regression coefficients, but since traditional dummy codes aren’t orthogonal, it messes things up when you’re just trying to partition variance (i.e. an ANOVA)."

"For an ANOVA, you should set your factors to use effects coding, rather than relying on the default treatment codes. You can do that with the contr.sum() function"

```{r contrasts}
# set effects contrasts (treat is an ordered factor w/ 4 levels, half with 2-lvls)
options(contrasts = c("contr.sum","contr.poly")) 
```

## Model 1: root growth in reponse to N-heterogeneity

```{r m1_root_growth}
plants_long.root <- plants_long[complete.cases(plants_long[ , c("root","nod"),]),]
plants_long.root <- droplevels(plants_long.root)
# str(plants_long.root)
## 638 obs.

# determine best probability distribution for data
ggplot(plants_long.root, aes(x = root.t)) + geom_density() + facet_grid(treat~half)
## right-skewed

require(fitdistrplus)
require(logspline)

descdist(plants_long.root$root, discrete = FALSE)
## exponential, beta, gamma
fit.gamma <- fitdist(plants_long.root$root.t, "gamma")
# fit.beta <- fitdist(plants_long.root$root.t, "beta") # has to be 0-1
fit.exp <- fitdist(plants_long.root$root.t, "exp")
plot(fit.gamma)
plot(fit.exp)
fit.gamma$aic 
fit.exp$aic
## Gamma, exp (best to worst)

# model for half
m1 <- glmer(root ~ (nod + line) * (treat/half) +
               (1|batch/tray/plant), 
               family = Gamma(link = "sqrt"),
               data = plants_long.root,
               control=glmerControl(optimizer="nloptwrap",
                            optCtrl=list(maxfun=2e4)))
summary(m1)
Anova(m1, type=3)

# check diagonstics
require(DHARMa)

# residual diagnostics using DHARMa
## Step 1: scaled residuals
simOut_m1 <- simulateResiduals(fittedModel = m1, n = 1000)
plot(simOut_m1) 
testUniformity(simOut_m1)
testDispersion(simOut_m1) 

# plot scaled residuals over each predictor (make sure their are even)
plotResiduals(plants_long.root$nod,  simOut_m1$scaledResiduals)

# lsmeans
require(emmeans)
require(lattice)
require(multcomp)
require(multcompView)

# treat:half interaction
HT_root.lsm <- emmeans(m1, ~ half | treat)
sum.HT_root.lsm <- summary(HT_root.lsm, type="response", infer= c(TRUE,TRUE), adjust = "bon")
save(sum.HT_root.lsm, file = "./models_files/sum.HT_root.lsm.Rdata")
(cld_HT_root.lsm <- cld(regrid(HT_root.lsm)))
# sig. diff at 20:80 and 2:98
plot(cld_HT_root.lsm)

# Dunnett-style contrasts (compare to control)
treat_con <- contrast(regrid(HT_root.lsm), method = "trt.vs.ctrl1")
con.HT_root.lsm <- summary(treat_con)
plot(treat_con)

# line:treat:half
LHT_root.lsm <- emmeans(m1, ~ half | treat + line)
sum.LHT_root.lsm <- summary(LHT_root.lsm, type="response", infer= c(TRUE,TRUE), adjust = "bon")
save(sum.LHT_root.lsm, file = "./models_files/sum.LHT_root.lsm.Rdata")

# Dunnett-style contrasts (compare to control)
treat_con <- contrast(regrid(LHT_root.lsm), method = "trt.vs.ctrl1")
con.LHT_root.lsm <- summary(treat_con, type="response")

lsmip(m1, half ~ treat, ylab = "Observed root biomass (mg)")
lsmip(m1, half ~ treat | line, ylab = "Observed root biomass (mg)")
```

## Model 2: nodulation in reponse to N-heterogeneity

Note: leave in any lower order terms, significant or not, for any higher order terms in the model. That means you have to leave in all non-significant two-way interactions for any significant 3-ways.

```{r m2_nodulation}
plants_long.nod <- plants_long[complete.cases(plants_long[ , c("nod","root"),]),]
plants_long.nod <- droplevels(plants_long.nod)
# str(plants_long.nod)
## 638 obs., 337 plants

## determine best probability distribution for data
ggplot(plants_long.nod, aes(x = nod)) + geom_density() + facet_grid(treat~half)
# right-skewed

require(fitdistrplus)
require(logspline)

descdist(plants_long.nod$nod, discrete = TRUE)
# negative bin, poisson, normal
fit.poiss <- fitdist(plants_long.nod$nod, "pois")
fit.norm <- fitdist(plants_long.nod$nod, "norm")
fit.nbinom <- fitdist(plants_long.nod$nod, "nbinom")
plot(fit.norm)
plot(fit.nbinom)
plot(fit.poiss)
fit.poiss$aic #5199.737
fit.norm$aic #4049.737
fit.nbinom$aic #3572.256
# nbin, normal, pois (best to worst)

# observation-level effect
plants_long.nod$obs_effect <- 1:nrow(plants_long.nod) # account for overdispersion

# model for AON
m2 <- glmer(nod ~ (root + line) * (treat/half) + 
               (1|batch/tray/plant) +
               (1|obs_effect), 
               family = poisson(link="sqrt"),
               data = plants_long.nod,
               control=glmerControl(optimizer="nloptwrap",
                            optCtrl=list(maxfun=2e4)))
summary(m2)
Anova(m2, type=3)

# residual dianostics using DHARMa
## Step 1: scaled residuals
simOut_m2 <- simulateResiduals(fittedModel = m2, n = 1000)
plot(simOut_m2)
testUniformity(simOut_m2)
testDispersion(simOut_m2)

# plot scaled residuals over each predictor (make sure their are even)
plotResiduals(plants_long.nod$leaf,  simOut_m2$scaledResiduals)

# lsmeans
require(emmeans)
require(lattice)
require(multcompView)

# treat:half interaction
HT_nod.lsm <- emmeans(m2, ~ half | treat)
sum.HT_nod.lsm <- summary(HT_nod.lsm, type="response", infer= c(TRUE,TRUE), adjust = "bon")
save(sum.HT_nod.lsm, file = "./models_files/sum.HT_nod.lsm.Rdata")
(cld_HT_nod.lsm <- cld(regrid(HT_nod.lsm)))
cld_HT_nod.lsm
# sig. diff at all but control
plot(cld_HT_nod.lsm)

# Dunnett-style contrasts (compare to control)
treat_con <- contrast(regrid(HT_nod.lsm), method = "trt.vs.ctrl1")
con.HT_nod.lsm <- summary(treat_con)
plot(treat_con)

# line:treat:half
LHT_nod.lsm <- emmeans(m2, ~ half | treat + line)
sum.LHT_nod.lsm <- summary(LHT_nod.lsm, type="response", infer= c(TRUE,TRUE), adjust = "bon")
save(sum.LHT_nod.lsm, file = "./models_files/sum.LHT_nod.lsm.Rdata")

# Dunnett-style contrasts (compare to control)
treat_con <- contrast(regrid(LHT_nod.lsm), method = "trt.vs.ctrl1")
con.LHT_nod.lsm <- summary(treat_con, type="response")

lsmip(m2, half ~ treat, ylab = "Observed nod biomass (mg)")
lsmip(m2, half ~ treat | line, ylab = "Observed nod biomass (mg)")
```

## Model 3: partner choice

```{r m3_partner_choice}
plants_long_strain.nod <- plants_long_strain[complete.cases(plants_long_strain[ ,
                                                    c("nod","root"),]),]
plants_long_strain.nod <- droplevels(plants_long_strain.nod)
# str(plants_long_strain.nod)
## 844 obs., 295 plants

# observation-level effect
plants_long_strain.nod$obs_effect <- 1:nrow(plants_long_strain.nod) # account for overdispersion

# model
m3 <- glmer(nod ~ (root + line) * (treat/half) * strain + 
               (1|batch/tray/plant) +
               (1|obs_effect), 
               family = poisson(link = "sqrt"),
               data = plants_long_strain.nod,
               control=glmerControl(optimizer="nloptwrap",
                            optCtrl=list(maxfun=2e4)))
summary(m3)
Anova(m3, type=3)

# residual dianostics using DHARMa
## Step 1: scaled residuals
simOut_m3 <- simulateResiduals(fittedModel = m3, n = 1000)
plot(simOut_m3) 
testUniformity(simOut_m3) 
testDispersion(simOut_m3)

# plot scaled residuals over each predictor (make sure they are even)
plotResiduals(plants_long_strain.nod$root,  simOut_m3$scaledResiduals)

## Check for singularities in model
tt <- getME(m3,"theta")
ll <- getME(m3,"lower")
min(tt[ll==0])
# issue: plant, batch, tray do not explain var.

# lsmeans
require(emmeans)
require(lattice)
require(multcompView)

# treat:half interaction
SHT_nod.lsm <- emmeans(m3, ~ half | treat + strain)
sum.SHT_nod.lsm <- summary(SHT_nod.lsm, type="response", infer= c(TRUE,TRUE), adjust = "bon")
save(sum.SHT_nod.lsm, file = "./models_files/sum.SHT_nod.lsm.Rdata")
(cld_SHT_nod.lsm <- cld(regrid(SHT_nod.lsm)))
# sig. diff at all but control
plot(cld_HT_nod.lsm)

# Dunnett-style contrasts (compare to control)
treat_con <- contrast(regrid(SHT_nod.lsm), method = "trt.vs.ctrl1")
con.SHT_nod.lsm <- summary(treat_con)
plot(treat_con)
```

## Whole-plant models

### mS1a: Shoot biomass

```{r mS1a_shoot}
# model
mS1a <- glmer(shoot ~ treat*line + 
                       (1|batch/tray), 
                     family=Gamma(link="sqrt"), 
                   data=plants)
plotresid(mS1a, shapiro = TRUE) # non-normal, no pattern
summary(mS1a)
Anova(mS1a, type=3)
summary(glht(mS1a, mcp(line="Tukey"))) # post-hoc comparisons
```

### mS1b: Root biomass

```{r mS1b_root}
# model total root biomass
mS1b <- glmer(tot_root ~ treat*line +
               (1|batch/tray), 
               family = Gamma(link="sqrt"),
               data = plants)
# check diagonstics
plotresid(mS1b, shapiro = TRUE)
summary(mS1b)
Anova(mS1b, type=3)
```

### mS1c: Nodules

```{r mS1c_nods}
# model total nodule number
plants$obs_effect <- 1:nrow(plants) # account for overdispersion
mS1c <- glmer(tot_nod ~ treat*line +
               (1|batch/tray) + 
                 (1|obs_effect), 
               family = poisson(link="sqrt"),
               data = plants)
# check diagonstics
plotresid(mS1c, shapiro = TRUE)
overdisp_fun(mS1c)
summary(mS1c)
Anova(mS1c, type=3)
```

### mS1d: Partner choice

```{r mS1d_pc}
# calculate whole-plant differences in preference
plants$tot_1022 <- plants$high_1022 + plants$low_1022
plants$pref <- (plants$tot_1022)/plants$tot_nod

# model
mS1d <- lmer(pref ~ treat*line + 
                   (1|batch/tray), 
                   data=plants)
plotresid(mS1d, shapiro = TRUE)
summary(mS1d)
Anova(mS1d, type=3)
```